---
title: "Term project"
author: "Micah Villarreal and Liz Lawler"
date: "5/6/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#NOTE TO USER: CHANGE TWO FILE PATHS (line 29 and 44) BEFORE PROCEEDING
```

```{r include=FALSE, message=FALSE}
if (!require(reshape2)) install.packages("reshape2",dependencies=TRUE)
if (!require(doBy)) install.packages("doBy",dependencies=TRUE)
if (!require(ggplot2)) install.packages("ggplot2",dependencies=TRUE)
if (!require(dplyr)) install.packages("dplyr",dependencies=TRUE)
if (!require(noncensus)) install.packages("noncensus",dependencies=TRUE)
if (!require(tidyr)) install.packages("tidyr",dependencies=TRUE)
require(reshape2)
require(doBy)
require(ggplot2)
require(dplyr)
require(noncensus)
require(tidyr)
```

```{r include = FALSE}
pha <- read.csv("~/Desktop/Term project/PHA_2014.csv"); head(pha) 
pha[pha == -1] <- NA
pha[pha == -4] <- NA
pha[pha == -5] <- NA
hcv <- filter(pha, pha$program == 3)
data(states)
states_fin <- states[, c(1, 3, 4)]
hcv <- left_join(hcv, states_fin, by = "state")
hcv$region <- as.character(hcv$region)
hcv$region[is.na(hcv$region)] <- "Island"
hcv$region <- as.factor(hcv$region)
table(hcv$region)
```

```{r, include = FALSE}
salaries <- read.csv("~/Desktop/Term project/2014EXEC_COMP.csv"); head(salaries)
strip_dol <- function(x) as.numeric((gsub("\\,", "", gsub("\\$", "", x))))
colnames(salaries)
salaries[,4:10] <- sapply(salaries[,4:10], strip_dol)
salaries_max <- salaries %>% group_by(PHA.Code) %>% top_n(1, Total.Compensation) %>% 
  distinct(salaries, PHA.Code, Total.Compensation, .keep_all = TRUE)
salaries_max$Total.Compensation[salaries_max$Total.Compensation == 0] <- NA 
salaries_max <- mutate(salaries_max, receive_bonus = Bonus > 0)
salaries_max$receive_bonus[is.na(salaries_max$Total.Compensation)] <- NA
salaries_max <- select(salaries_max, Total.Compensation, code = PHA.Code, receive_bonus)
hcv <- left_join(hcv, salaries_max, by = "code")
hcv <- mutate(hcv, num_hh = total_units * (pct_occupied/100))
hcv <- mutate(hcv, poverty_area = tpoverty >=20)
hcv <- mutate(hcv, num_fem = total_units * (pct_occupied/100) * (pct_female_head/100))
hcv <- mutate(hcv, num_male = total_units * (pct_occupied/100) * (1-(pct_female_head/100)))
hcv <- mutate(hcv, total_rent  = rent_per_month + spending_per_month)
hcv <- mutate(hcv, income_monthly  = hh_income/12)
hcv <- mutate(hcv, rent_burden  = total_rent/income_monthly)
```

## Introduction
< insert brief intro about what we did (from our one-page handout) >

### Households Receiving Subsidized Housing by Region
We first decided to look at the number of households by census region, and display the porportion of female heads of household to male heads of household. As you can see, across all regions, the majority of households receiving subsidized housing are headed by females.

```{r echo=FALSE}
hcv_collapse <- summaryBy(num_male+num_fem~region,data=hcv,FUN=sum,na.rm=TRUE)
hcv_collapse_melt <-melt(hcv_collapse,id.var="region")

ggplot(data = hcv_collapse_melt, aes(x = reorder(region, -value), y = value, 
  fill = factor(variable))) + geom_bar(stat="identity") + theme_bw() + labs(x="Region")  +
  ggtitle("Households Receiving Subsidized Housing\n by Region") +
  scale_y_continuous(name="Count",labels=scales::comma) +
  theme(legend.position="bottom",legend.direction = "horizontal", legend.title=element_blank()) +
  scale_fill_manual(values=c("steelblue1", "lightpink"),labels=c("Male-Headed", "Female-Headed")) +
  theme(plot.title = element_text(hjust = 0.5))
```

### Total Compensation of Public Housing Authority Executives
Next, we looked at the distribution of the maximum Total Compensation <will write story after putting in all graphics>
```{r echo=FALSE}
hist(hcv$Total.Compensation/1000, breaks = "FD", 
     col=rgb(0.1,0.5,0.8,0.5), main = "Earnings of the Top-Paid Employee at\n every Public Housing Authority",
     xlab = "Total Compensation (in thousands of dollars)", ylab = "Number of HCV Programs",las=1,xaxt="n",border=F,
     xlim = c(0, 260), ylim=c(0,300))
axis(side=1, at=seq(0,260,10))
```
### Average Rent Burden of Public Housing Authority Tenants
Next, we ....
```{r echo = FALSE}
hist(hcv$rent_burden, breaks = "FD", 
     col=rgb(0.8,0.3,0.6,0.5), xlim = c(0.5, 2.5), ylim=c(0,300),
     main = "Average Rent Burden",
     xlab = "Ratio of Pre-Subsidy Rent to Household Monthly Income", ylab = "Number of HCV Programs",border=F) 
```
### Average Months Waiting for a Housing Choice Voucher
We then...
```{r echo = FALSE}
hist(hcv$months_waiting, breaks = "FD", prob = TRUE,
     ylim = c(0,.04),
     col=rgb(0.2,0.8,0.5,0.5),border=F,
     main="Time Spent Waiting for a Home",
     xlab="Average Months",ylab="Proportion of HCV Programs",xaxt="n")
axis(side=1, at=seq(0,200,10))
a <- 1/mean(hcv$months_waiting, na.rm = TRUE)
curve(dexp(x, a), add = TRUE)
```
It makes sense that an exponential curve fits "number of months waiting" very well. This curve arises when examining the length of time between events in Poisson processes. 
