---
title: "Final project"
author: "Liz Lawler and Micah Villarreal"
date: "5/6/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
#NOTE TO USER: CHANGE TWO FILE PATHS (line 29 and 44) BEFORE PROCEEDING
```

```{r include=FALSE, message=FALSE}
if (!require(reshape2)) install.packages("reshape2",dependencies=TRUE)
if (!require(doBy)) install.packages("doBy",dependencies=TRUE)
if (!require(ggplot2)) install.packages("ggplot2",dependencies=TRUE)
if (!require(dplyr)) install.packages("dplyr",dependencies=TRUE)
if (!require(noncensus)) install.packages("noncensus",dependencies=TRUE)
if (!require(tidyr)) install.packages("tidyr",dependencies=TRUE)
if (!require(knitr)) install.packages("knitr",dependencies=TRUE)
if (!require(kableExtra)) install.packages("kableExtra",dependencies=TRUE)
if (!require(apa)) install.packages("apa",dependencies=TRUE)
require(reshape2)
require(doBy)
require(ggplot2)
require(dplyr)
require(noncensus)
require(tidyr)
require(knitr)
require(kableExtra)
require(apa)
```

```{r include = FALSE}
pha <- read.csv("~/Desktop/Term project/PHA_2014.csv"); head(pha) 
pha[pha == -1] <- NA
pha[pha == -4] <- NA
pha[pha == -5] <- NA
hcv <- filter(pha, pha$program == 3)
data(states)
states_fin <- states[, c(1, 3, 4)]
hcv <- left_join(hcv, states_fin, by = "state")
hcv$region <- as.character(hcv$region)
hcv$region[is.na(hcv$region)] <- "Island"
hcv$region <- as.factor(hcv$region)
table(hcv$region)
```

```{r, include = FALSE}
salaries <- read.csv("~/Desktop/Term project/2014EXEC_COMP.csv"); head(salaries)
strip_dol <- function(x) as.numeric((gsub("\\,", "", gsub("\\$", "", x))))
colnames(salaries)
salaries[,4:10] <- sapply(salaries[,4:10], strip_dol)
salaries_max <- salaries %>% group_by(PHA.Code) %>% top_n(1, Total.Compensation) %>% 
  distinct(salaries, PHA.Code, Total.Compensation, .keep_all = TRUE)
salaries_max$Total.Compensation[salaries_max$Total.Compensation == 0] <- NA 
salaries_max <- mutate(salaries_max, receive_bonus = Bonus > 0)
salaries_max$receive_bonus[is.na(salaries_max$Total.Compensation)] <- NA
salaries_max <- select(salaries_max, Total.Compensation, code = PHA.Code, receive_bonus)
hcv <- left_join(hcv, salaries_max, by = "code")
hcv <- mutate(hcv, poverty_area = tpoverty >=20)
hcv <- mutate(hcv, num_hh = total_units * (pct_occupied/100))
hcv <- mutate(hcv, num_fem = total_units * (pct_occupied/100) * (pct_female_head/100))
hcv <- mutate(hcv, num_mother = total_units * (pct_occupied/100) * (pct_female_head_child/100))
hcv <- mutate(hcv, num_male = total_units * (pct_occupied/100) * (1-(pct_female_head/100)))
hcv <- mutate(hcv, total_rent  = rent_per_month + spending_per_month)
hcv <- mutate(hcv, income_monthly  = hh_income/12)
hcv <- mutate(hcv, rent_burden  = total_rent/income_monthly)
```

# Introduction
For our term project, we examined relationships found within two datasets.  
1. PHA-“program type”-level  data about the state of subsidized housing in the United States in 2014, courtesy of the Department of Housing and Urban Development (HUD). Though HUD engages in several different housing assistance program types, we limited our analysis to Housing Choice Voucher (HCV) programs.  https://www.huduser.gov/portal/datasets/assthsg.html  
2. PHA-person-level data about the compensation (salary, bonus and total compensation) of the top three paid employees at every public housing authority in the United States. We limited this data to the top paid individual at each PHA. https://www.hud.gov/program_offices/public_indian_housing/programs/pha_executive_compensation  
<br>We constructed a PHA-level dataset with programmatic, geographic and compensation information about every HCV program in the United States, as well as rich demographic data about each program’s clients.
 

## Households Receiving Subsidized Housing by Region
We first decided to look at the number of households by census region, and display the porportion of female heads of household to male heads of household. As you can see, across all regions, the majority of households receiving subsidized housing are headed by females. We were surprised to see how many households were headed by females *without* children. Further exploration could include examining the age distribution within these households; one hypothesis might be that the regions with households headed by females with no children also have a higher percentage of elderly people in public housing. 

```{r echo=FALSE}
hcv_collapse <- summaryBy(num_male+num_fem+num_mother~region,data=hcv,FUN=sum,na.rm=TRUE)
hcv_collapse <- mutate(hcv_collapse, num_nonmother  = num_fem.sum-num_mother.sum)
hcv_collapse <- select(hcv_collapse, region,num_male.sum, num_nonmother,num_mother.sum)
hcv_collapse_melt <-melt(hcv_collapse,id.var="region")
ggplot(data = hcv_collapse_melt, aes(x = reorder(region, -value), y = value, 
                                     fill = factor(variable))) + geom_bar(stat="identity") + theme_bw() + labs(x="Region")  +
  ggtitle("Households Receiving Subsidized Housing\n by Region") +
  scale_y_continuous(name="Count",labels=scales::comma) +
  theme(legend.position="bottom",legend.direction = "horizontal", legend.title=element_blank()) +
  scale_fill_manual(values=c("mediumseagreen", "lightsalmon1","orchid"),
                    labels=c("Headed\n by a Man", "Headed by a\n Woman, No Children","Headed by a\n Woman with Children")) +
  theme(plot.title = element_text(hjust = 0.5))

```

## Total Compensation of Public Housing Authority Executives  
### Distribution
Next, we looked at the distribution of the maximum Total Compensation.
```{r echo=FALSE}
hist(hcv$Total.Compensation/1000, breaks = "FD", 
     col=rgb(0.1,0.5,0.8,0.5), main = "Earnings of the Top-Paid Employee at\n every Public Housing Authority",
     xlab = "Total Compensation (in thousands of dollars)", ylab = "Number of HCV Programs",las=1,xaxt="n",border=F,
     xlim = c(0, 260), ylim=c(0,300))
axis(side=1, at=seq(0,260,10))
```

### Relationships and Comparisons
We then looked at the relationship between total compensation and poverty areas. We performed a permutation test on compensation of highest paid employee by whether the program was located in a high poverty area or not. 
```{r echo = FALSE, results = "asis", message = FALSE}
PoorInd <- which(hcv$poverty_area)
Total.Compensation <- hcv$Total.Compensation
Obs <- mean(Total.Compensation[PoorInd],na.rm=TRUE)-mean(Total.Compensation[-PoorInd],na.rm=TRUE)
```

* Observed difference in means: 
```{r echo = FALSE, results = "asis", message = FALSE}
Obs
N <-10000; diff <- numeric(N)
for (i in 1:N) {
  scramble <- sample(hcv$poverty_area, length(hcv$poverty_area))
  PoorInd <- which(scramble)
  diff[i] <- mean(Total.Compensation[PoorInd],na.rm=TRUE)-mean(Total.Compensation[-PoorInd],na.rm=TRUE)
}
```

* Permutation test p-value: 
```{r echo = FALSE, results = "asis", message = FALSE}
pvalue <- mean(diff > Obs); pvalue 
```

With a p-value of 0.001%, this difference is significant. However, we actually expected that being located in a high-poverty area would be associated with LOWER executive pay. It would make sense that if a lot of the people in a town fall under the poverty line, which is calculated at a national level, that cost of living in the area (rent, food, gas, etc) would be lower for everyone in the town. In addition, job opportunities are probably scarcer so wages are lower. We were surprised to find that executives in high poverty areas actually receive statistically significantly higher total compensation.  
<br>
We then explored the relationship between total compensation and racial minorities, as well as total comp and tenant income.<br>
**Total comp and racial minorities**
<br>
Positive correlation between Total Executive Compensation and Racial Minority:
```{r echo = FALSE, results = "asis", message = FALSE, warning = FALSE}
cor(hcv$Total.Compensation, hcv$pct_minority, use = "complete.obs")
```

```{r echo = FALSE, results = "asis", message = FALSE, warning = FALSE}
ggplot(hcv, aes(x=pct_minority, y=Total.Compensation, color=pct_minority)) + 
  scale_x_continuous(name="Percent of Racial Minority Clients") + geom_point()+
  scale_y_continuous(name="Largest 'Total Compensation' @ PHA",labels=scales::comma) +
  ggtitle("Race and PHA Executive Compensation") +
  geom_smooth(method = 'lm',color='black') + theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5))+scale_color_gradientn(colours = rainbow(5))
```
<br>We did not expect that for each additonal percentage point minority in the HCV Program, executive compensation increases by $414, and it's statistically significant (per our linear regression)! We expected these things to be unrelated.
<br>
**Total comp and tenant income**<br>
Positive correlation between Total Executive Compensation and Tenant Income:
```{r echo = FALSE, results = "asis", message = FALSE, warning = FALSE}
hcv_mainland <- filter(hcv, hcv$region != "Island")
cor(hcv_mainland$Total.Compensation, hcv_mainland$hh_income, use = "complete.obs")
```

```{r echo = FALSE, results = "asis", message = FALSE, warning = FALSE}
ggplot(hcv_mainland, aes(x=hh_income, y=Total.Compensation), 
       group=region) + 
  geom_point(aes(shape=region, color=region)) + 
  scale_x_continuous(name="Average Tenant Household Income",labels=scales::comma) +
  scale_y_continuous(name="Largest 'Total Compensation' @ PHA",labels=scales::comma) +
  ggtitle("Relationship between Tenant Income\nand PHA Executive Compensation") +
  stat_smooth(method = 'lm') + theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```
<br>As we expected, total compensation of PHA executives increases with tenant income.


## Average Rent Burden of Public Housing Authority Tenants
Next, we examined the distribution of “rent burden” or the ratio of average monthly household income to average monthly pre-subsidy rent. Pre-subsidy refers to the sum of the household contribution and the HCV-contribution.

```{r echo = FALSE}
hist(hcv$rent_burden, breaks = "FD", 
     col=rgb(0.8,0.3,0.6,0.5), xlim = c(0.5, 2.5), ylim=c(0,300),
     main = "Average Rent Burden",
     xlab = "Ratio of Pre-Subsidy Rent to Household Monthly Income", ylab = "Number of HCV Programs",border=F) 
```

## Average Months Waiting for a Housing Choice Voucher
### Distribution
Next we looked at xyz

```{r echo = FALSE}
hist(hcv$months_waiting, breaks = "FD", prob = TRUE,
     col=rgb(0.2,0.8,0.5,0.5),border=F,
     main="Time Spent Waiting for a Home",
     xlab="Average Months",ylab="Proportion of HCV Programs",xaxt="n")
axis(side=1, at=seq(0,200,10))
a <- 1/mean(hcv$months_waiting, na.rm = TRUE)
curve(dexp(x,a), from = 0, add = TRUE)
```

It makes sense that an exponential curve fits "number of months waiting" very well. This curve arises when examining the length of time between events in Poisson processes. 

### Relationship with poverty area
Next, we used two methods to examine the relationship between "average months waiting" and whether or not the respective area was considered a "poverty area" (Census Bureau designates census tracts with a poverty rate >=20% as "poverty areas").

First, we created a contingency table of "months waiting" categories crossed with "poverty areas"" (Census Bureau designates census tracts with a poverty rate >=20% as "poverty areas").

```{r echo = FALSE, results = "asis", message = FALSE}
attach(hcv)
hcv$mw_bin[hcv$months_waiting < 6] <- "0 to < 6 mo."
hcv$mw_bin[hcv$months_waiting >= 6  & hcv$months_waiting < 12] <- "6 to < 12 mo."
hcv$mw_bin[hcv$months_waiting >= 12  & hcv$months_waiting < 18] <- "12 to < 18 mo."
hcv$mw_bin[hcv$months_waiting >= 18  & hcv$months_waiting < 24] <- "18 to < 24 mo."
hcv$mw_bin[hcv$months_waiting >= 24  & hcv$months_waiting < 30] <- "24 to < 3 mo."
hcv$mw_bin[hcv$months_waiting >= 30  & hcv$months_waiting < 36] <- "30 to < 36 mo."
hcv$mw_bin[hcv$months_waiting >= 36] <- "36+ mo."
detach(hcv)
hcv$mw_bin <- factor(as.factor(hcv$mw_bin), levels = 
                       c("0 to < 6 mo.", "6 to < 12 mo.", "12 to < 18 mo.",
                         "18 to < 24 mo.", "24 to < 3 mo.", "30 to < 36 mo.",
                         "36+ mo."))
pov_mo_tbl <- table(hcv$mw_bin, hcv$poverty_area)
Expected <- outer(rowSums(pov_mo_tbl), colSums(pov_mo_tbl))/sum(pov_mo_tbl)
kable(pov_mo_tbl, "html", caption = "Observed Distribution of Average Months Waiting per Poor and Non-poor Areas") %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, position = "float_left") %>% add_header_above(c("Average Months Waiting" = 1, "Poverty Area" = 2))
kable(Expected, "html", caption = "Expected Distribution of Average Months Waiting per Poor and Non-poor Areas") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, position = "float_left") %>% add_header_above(c("Average Months Waiting" = 1, "Poverty Area" = 2))
```
<br><br>
Chi-square test:
```{r echo = FALSE, results = "asis", message = FALSE}
chisq_apa(chisq.test(hcv$mw_bin, hcv$poverty_area))
```

There is about a 1% chance that the observed contingency table arose by chance. Logically this makes sense. Much fewer HCV programs located in poorer places (poverty_area = TRUE) experience short wait times (0-6 months) than we would expect, but many more experience wait times in every other bucket than we would expect if wait time was random. Many more HCV programs than we would expect in non-poor areas experience short wait times, but many fewer experience wait times in every other bucket.
<br><br><br><br>
Then, we looked at the significane of the difference in mean average waiting time of poor areas vs non-poor areas by conducting a permutation test:
<br>
```{r echo = FALSE, message = FALSE}
PoorInd <- which(hcv$poverty_area)
mw <- hcv$months_waiting
obs_mw <- mean(mw[PoorInd], na.rm = TRUE) - mean(mw[-PoorInd], na.rm = TRUE)
```
* Observed difference in means: 
```{r echo = FALSE, results = "asis", message = FALSE}
obs_mw
N <- 10^5
diffs <- numeric(N)
for (i in 1:N) { 
  scramble <- sample(hcv$poverty_area, length(hcv$poverty_area), replace = FALSE)
  index <- which(scramble)
  diffs[i] <- mean(mw[index],na.rm=TRUE)-mean(mw[-index],na.rm=TRUE)
}
```

* Permutation test p-value: 
```{r echo = FALSE, results = "asis", message = FALSE}
pvalue <- (sum(diffs >= obs_mw)+1)/(N+1); pvalue
``````

<br>According to our permutation test, there's a 95% chance of the difference in mean waiting times between poor and non poor areas occurring by chance. This difference in mean is not significant.
<br><br><br>**need to add in how we can see two different results in this**
 
